---
title: "Class 9: Candy Mini-Project"
author: "Sean Rye (PID: A17793986)"
format: pdf
toc: TRUE
---

## Background 

In this mini-project, you will explore FiveThirtyEight's Halloween Candy dataset.

We will use lots of **ggplot** some basic stats, correlation, analysis and PCA to make sense of the landscape of US candy -

## Data Import

Our dataset is a CSV file 

```{r}
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```


> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Reese's Peanut Butter cup") |> 
  select(winpercent)
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```


> Q7. What do you think a zero and one represent for the candy$chocolate column?

Ones and Zeros represent whether the candy contains chocolate 

## Exploratory Analysis

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy, aes(winpercent)) +
  geom_histogram()
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values lack symmetry

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.candy <- candy[candy$chocolate == 1,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.win <- candy[candy$fruity == 1, ]$winpercent
mean(fruit.win)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
## Overall Candy Rankings

> Q13. What are the five least liked candy types in his set?

```{r}
y <- c("y", "a","z")
sort(y)
```
```{r}
y
order(y)
```

```{r}
ord.ind <- order(candy$winpercent)
head( candy[ord.ind,], 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail( candy[ord.ind,], 5)
```



> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, row.names(candy)) +
  geom_col() + 
  ylab("")
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(row.names(candy), winpercent)) +
  geom_col() + 
  ylab("")
```
```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(row.names(candy), winpercent),
      fill=chocolate) +
  geom_col() + 
  ylab("")
```

We need a custom color vector

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[candy$chocolate==1] <- "chocolate"
my_cols[candy$bar==1] <- "brown"
my_cols[candy$fruity==1] <- "darkgreen"
my_cols
```

```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(row.names(candy), winpercent),
      fill=chocolate) +
  geom_col( fill=my_cols) + 
  ylab("")
```
> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Taking a look at Pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, candy$pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12)], n=5 )
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Exploring the correlation structures

```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)

corrplot(cij)
```
## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Score Plot...

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label=row.names(pca$x)) + 
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps = 7, size=3.3,col=my_cols)
```


```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The orginial variables picked up strongly by PC1 in the positive direction include fruity, pluribus, and hard which can be attributed to their shared positive correlations with each other

```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
ggplot(pca$rotation) + 
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1) ) + 
  geom_col()
```

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on the exploratory analysis, correlation findings, and PCA results, the combination of characteristics yielding a "winning" candy is 'chocolate' and 'bar' as these characteristics showed high winning percentages and a positive correlation with each other.
