---
title: "Class 12 HW"
author: "Sean Rye (PID: A17793986) "
format: pdf
toc: true
---

## Retrieving File 

```{r}
genotype.df <- read.table("Genotype.csv", row.names = 1)
genotype.df
```

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

Determine the # of samples for each genotype
```{r}
library(dplyr)

genotype.df %>%
  count(geno)
```


Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

## Generate an intial boxplot

```{r}
library(ggplot2)

ggplot(genotype.df, aes(x = geno, y = exp)) +
  geom_boxplot()
```

## Add some color to the plots

```{r}
ggplot(genotype.df, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot()
```

Based on the boxplots, we can infer that the "A" allele has the higher expression with the median value being the highest amongst the three genotypes for A|A while G|G has the lowest expression. The SNP seems to affect expression as the genotypes containing the "A" allele shows a higher expression, meaning that the SNP doesn't affect ORMDL3 expression.


