---
title: "Class 5: Data Viz with ggplot"
author: "Sean Rye (PID: A17793986)"
format: pdf
toc: true
---

## Background

There are lot's of ways to make figures in R. These include the so-called "base r" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command 
`install-packages()`. 

> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want!

Everytime we want to use one of these "add-on" packages we need to load it up in R with `library` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The **geom**etry, the type of plot 

```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) + 
  labs(title="Stopping distance of old cars", subtitle = "Data from the `cars` object",  x="Speed (mph)", y="Distance(ft)")
```

Add a line to better show the relationship between speed and distance


render it out
```{r}
p + theme_bw()
```

## Gene expression plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version of the plot

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) + 
  geom_point()
```

```{r}
table(genes$State)
```

Version 2 let's color by `State` so we can see the "up" and "down" significant genes to all the unchanging genes

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point()
```
Version 3 plot, let's modify the default colors to something we like

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() + 
  scale_colour_manual(values=c("darkgreen", "purple", "navy")) + 
  
  labs(x="Control (no drug)", y="Drug", title ="Gene Expression Changes upon GLP-1 drug")
```

## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(data=gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point(alpha=0.3)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above.

```{r}
ggplot(data=gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point(alpha=0.3) + 
  facet_wrap(~continent)
```

## Custom plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up...


```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
filter(gapminder, year==1977, country=="Ireland")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977, 2007))
head(input)
```

```{r}
gapminder_1977 <- gapminder %>% filter(year==1977 | year==2007)

ggplot(gapminder_1977) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent), alpha=0.7) + 
  scale_size_area(max_size = 10) + 
  facet_wrap(~year)
```

