---
title: "Class 7: Machine Learning 1"
author: "Sean Rye (PID: A17793986)"
format: pdf
toc: true
---

## Background

Today we will begin our exploration of important machine learning methods with a focus on **clustering** and **dimensionality reduction**.  

To start testing thse methods let's make up some sample dara to cluster where we know what the answer should be.


```{r}
hist( rnorm(3000, mean = 10) )
```
> Q. Can you generate 30 numbers centered at +3 and 30 numbers at -3 taken at random from a normal distribution?

```{r}
tmp <- c(rnorm(30, mean = 3),
          rnorm(30, mean= -3) )

x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

## K-means clustering 

The main function in "base R" for K-means clustering is called `kmeans()`, let's try it out:

```{r}
k <- kmeans(x, centers=2)
k 
```
> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

> Q. What component of your kmeans result object has the cluster size (i.e. how many points are in each cluster?

```{r}
k$size
```

> Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster?)

```{r}
k$cluster
```

> Q. Plot the result of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col=c(1,2))
points(k$centers, col="blue", pch=15, cex=2)
```

> Q. Can you run `kmeans()` again and cluster `x` into 4 clusters and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k4 <- kmeans(x, centers=4)
plot(x,col=k4$cluster)
points(k4$centers, col="blue", pch =15, cex=2)
```

> **Key point:** Kmeans will always return the clustering that we ask for (this the "K" or "centers" in K-means)

```{r}
k$tot.withinss
```

## Hierarchical Clustering 

the main function for hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you can not just pass your input data directly to `hclust()` - it needs a "distance matrix" as input. we can get this from lot's of places including the `dist()` function

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut" the dendrogram or "tree" at a given height to yield our "clusters". For thi s we use the function `cutree()`

```{r}
plot(hc)
abline(h=10, col="red")
grps <- cutree(hc, h=10)
```

```{r}
grps
```

> Q. Plot our data `x` colored by the clusterinv result from `hclust()` and `cutree()`?

```{r}
grps <- cutree(hc, h = 10)
plot(x, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics.

### PCA of UK food data 

Read data on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

It looks like the row names are not set properly. We cann fix this 

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```

A better way to do this is to fix the row names assignment at the import time:

```{r}
x <- read.csv(url, row.names=1)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
nrow(x)
ncol(x)
x <- read.csv(url, row.names=1)
head(x)
```

There are 17 rows and 5 columns

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

I prefer the second method of altering the original command rather than the first method 

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

Changing the "beside" setting from true to false results in the above plot being created.


> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

### Heatmap 

We can install the **pheatmap** package with the `install.packages()` command that we used previously. Remember that we always run this in the console and not a code chunk1 in our quarto document

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

Of all these plot really only the `pairs()` plot was useful. This however took a bit of work to interpret and will not scale when I am looking at much bigger datasets

## PCA the rescue

The main function in "base R" for PCA is called `prcomp()` 

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

> Q. How much variance is captured in the first PC?

67.4%

> Q. How many PCs do I need to capture at least 90% of the total variance in the dataset?

Two PCs capture 96.5% of the total variance. 

> Q. Plot our main PCA results. Folks can call this different things depending on their field of study e.g. "PC plot", "ordienation plot", "Score plot", "PC1 vs PC2 plot"...

```{r}
attributes(pca)
```
To generate our PCA score plot we want `pca$x` component of the result object

```{r}
pca$x
```

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col = my_cols, pch = 16)
```

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col = my_cols)
```

## Digging deeper (variable loadings)

How do the original variables (i.e. the 17 different foods) contribute 

